---
title: "ivamse: Choosing the Dictionary and Penalty for IV-LASSO"
author: "Yukun Ma, Manu Navjeevan, Bohdan Salahub"
output:
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{ivamse: Choosing the Dictionary and Penalty for IV-LASSO}
  %\VignetteDepends{ivamse}
  %\VignetteKeywords{instrumental variables, lasso, model selection}
  %\VignettePackage{ivamse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r preliminaries, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 3.6, dpi = 110)
set.seed(20260904)
```

## Overview

Consider estimating

$$y_i = \beta x_i + w_i'\gamma + \varepsilon_i, \qquad x_i = \Pi_i + v_i,$$

where $x_i$ is endogenous, $w_i$ collects included exogenous controls, and many
instruments are available. Fitting the first stage by the lasso requires two
choices: which dictionary of technical instruments to use, and how heavily to
penalize.

First-order asymptotics do not distinguish the candidates, however, since every
sufficiently good first stage gives

$$\sqrt{n}(\hat\beta - \beta) = \frac{\sqrt{n}\,\mathbb{E}_n[\Pi_i\varepsilon_i]}{H} + o_p(1),$$

whose leading term does not involve the first-stage estimate. Separating the
candidates therefore requires the terms that this expansion discards. The
criterion implemented in `ivamse()` estimates those terms and ranks the
candidates by them.

```{r setup}
library("ivamse")
```

## Illustration: A sparse first stage

The four leading instruments enter the first stage, the remaining fifty-six do
not, and the structural and first-stage errors are correlated, so the regressor
is endogenous.

```{r data}
make_data <- function(N = 800, p = 60, rho = 0.7) {
  z <- matrix(rnorm(N * p), N, p)
  colnames(z) <- paste0("z", seq_len(p))
  v <- rnorm(N)
  e <- rho * v + sqrt(1 - rho^2) * rnorm(N)
  w <- rnorm(N)
  x <- z[, 1] + 0.8 * z[, 2] + 0.6 * z[, 3] + 0.4 * z[, 4] + w + v
  y <- 2 * x + w + e
  data.frame(y, x, w, firm = factor(rep_len(1:80, N)), z)
}

d <- make_data()
instruments <- reformulate(paste0("z", 1:60))
```

The true coefficient on `x` is 2. Ordinary least squares ignores the
correlation between the two errors:

```{r ols}
m_ols <- lm(y ~ x + w, data = d)
coef(m_ols)["x"]
```

Thus, OLS yields an estimate of `r round(coef(m_ols)["x"], 3)`.

## Fitting the model

The model formula has three parts on the right-hand side, giving the exogenous
regressors, the endogenous regressor, and the excluded instruments. When
`dictionaries` is supplied, as it is here, the third part is ignored and the
dictionaries take its place.

```{r fit}
m_iv <- ivamse(y ~ w | x | z1, dictionaries = list(all = instruments), data = d)
summary(m_iv)
```

The summary reports the selected dictionary and penalty, the effective
dimension of the corresponding first stage, and the two estimated error
moments. The implied correlation is attenuated relative to the `rho` of the
design, because it is computed from a preliminary estimate of $\beta$ rather
than from the truth; Ma, Navjeevan and Salahub analyze the effect of that
preliminary estimate on the criterion.

## The candidate grid

`plot()` shows the criterion and the effective dimension across the penalty
grid, with the selected candidate marked.

```{r plot}
plot(m_iv)
```

The same information is held in the candidate table. The multiplier `kappa`
expresses each penalty as a fraction of the plug-in penalty of Belloni, Chen,
Chernozhukov and Hansen, and `implied_alpha` is positive for those candidates
that satisfy the penalty condition of Ma, Navjeevan and Salahub.

```{r candidates}
cand <- m_iv$candidates
cand[seq(1, nrow(cand), by = 4),
     c("lambda", "kappa", "h", "d", "criterion", "implied_alpha")]
```

`selected()` reports which instruments the chosen first stage uses.

```{r selected}
selected(m_iv)$instruments
```

## Endogeneity and the selected dimension

The bias term of the criterion is weighted by the squared covariance of the two
errors, so a more endogenous regressor should be met with heavier
regularization and a simpler first stage. The specification is identical in
every row below; only `rho` and the draw of the data change.

```{r endogeneity}
sweep_rho <- function(rho) {
  fit <- ivamse(y ~ w | x | z1, dictionaries = list(all = instruments),
                data = make_data(rho = rho))
  c(rho = rho, estimate = unname(coef(fit)[1]), dimension = fit$selected$d)
}
t(sapply(c(0.1, 0.4, 0.7, 0.9), sweep_rho))
```

The effective dimension falls sharply as the correlation rises, in the
direction the criterion predicts. A single draw at each correlation is an
illustration rather than evidence; Ma, Navjeevan and Salahub report the
corresponding Monte Carlo study.

## Selection rules

Cross-validation ranks candidates by first-stage fit alone, and the plug-in rule
targets a penalty that dominates first-stage noise. Neither uses the outcome
equation, so neither responds to the endogeneity of the regressor.

```{r rules}
compare <- function(rule) {
  fit <- ivamse(y ~ w | x | z1, dictionaries = list(all = instruments),
                select = rule, data = d)
  c(estimate = unname(coef(fit)[1]), dimension = fit$selected$d)
}
data.frame(rule = c("amse", "cv", "bcch"),
           t(sapply(c("amse", "cv", "bcch"), compare)))
```

Cross-validation selects a considerably richer first stage than the criterion
does at this correlation.

## Several dictionaries

Ma, Navjeevan and Salahub select over the dictionary as well as the penalty.
A named list of one-sided formulas supplies the candidate dictionaries, and the
criterion is on a common scale, so candidates from different dictionaries are
directly comparable.

```{r dictionaries}
m_dict <- ivamse(
  y ~ w | x | z1,
  dictionaries = list(
    small  = ~ z1 + z2 + z3,
    medium = ~ z1 + z2 + z3 + z4 + z5 + z6 + z7 + z8,
    wide   = ~ (z1 + z2 + z3 + z4) * (z5 + z6 + z7 + z8)
  ),
  data = d
)
m_dict$selected$dictionary
plot(m_dict)
```

Ties are broken toward the earlier candidate, so the order of the list matters
and the simpler dictionaries should be listed first.

## Controls and clustered standard errors

Included exogenous controls enter both equations and are partialled out. They
are never penalized, and the effective sample size becomes $n = N - q$, where
$q$ is the rank of the controls.

Clustering affects the reported standard errors only. Selection always uses the
homoscedastic criterion, since that is what the theory covers.

```{r cluster}
m_cl <- ivamse(y ~ w | x | z1, dictionaries = list(all = instruments),
               cluster = ~ firm, data = d)
confint(m_cl)
```

Because `bread()` and `estfun()` methods are provided, the variance estimators
of the **sandwich** package apply directly:

```{r sandwich, eval = requireNamespace("sandwich", quietly = TRUE)}
sqrt(diag(sandwich::vcovHC(m_cl, type = "HC1")))
```

## Limitations

The model may contain only one endogenous regressor; the theory does not cover
a vector $x_i$.

The criterion contains an offset common to every candidate, so its level is not
an estimate of mean squared error and only differences across candidates are
meaningful.

Standard errors treat the selected first stage as fixed. They account for
neither first-stage estimation error nor the selection step, and they are not
valid under weak identification.

The criterion is derived under homoscedastic Gaussian errors. Simulations with
Laplace errors leave the ranking of candidates essentially unchanged, but
heteroscedastic selection is not covered by any result.

## References

Belloni, A., Chen, D., Chernozhukov, V., and Hansen, C. (2012). Sparse models
and methods for optimal instruments with an application to eminent domain.
*Econometrica* **80**, 2369--2429.

Donald, S. G., and Newey, W. K. (2001). Choosing the number of instruments.
*Econometrica* **69**, 1161--1191.

Ma, Y., Navjeevan, M., and Salahub, B. Choosing the dictionary and penalty for
IV-LASSO.
